---
title: "CellDynamicST: Full Pipeline Analysis Notebook"
author: "CellDynamicST Team"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    self-contained: true
params:
  config_path: "experiment_config.yaml"
  output_dir: "cdst_output"
  n_cores: 4
  steps: "all"
execute:
  warning: false
  message: false
---

## About This Notebook

This parameterized Quarto notebook runs the complete CellDynamicST pipeline on
your spatial transcriptomic data. To use it with your own data, edit the `params`
block above to point to your configuration file and desired output directory,
then render the notebook.

```bash
# Render from the command line
quarto render 01_full_pipeline.qmd \
  -P config_path:my_config.yaml \
  -P output_dir:my_output \
  -P n_cores:8
```

## Setup

```{r setup}
library(CellDynamicST)
library(ggplot2)
library(dplyr)

config_path <- params$config_path
output_dir <- params$output_dir
n_cores <- params$n_cores

cat("Configuration:", config_path, "\n")
cat("Output directory:", output_dir, "\n")
cat("Cores:", n_cores, "\n")
```

## Load Configuration

```{r config}
config <- cdst_load_config(config_path)
print(config)
```

## Step 1: Data Registration

```{r register}
samples <- cdst_register_samples(config)
merged <- cdst_merge_samples(samples, config)
cat("Total cells after merge:", ncol(merged), "\n")
```

## Step 2: Quality Control & Normalization

```{r preprocess}
merged <- cdst_qc_filter(merged, config)
cat("Cells after QC:", ncol(merged), "\n")

merged <- cdst_normalize(merged, config)
```

### QC Visualization

```{r qc-plot, fig.width=15, fig.height=5}
cdst_plot_qc(merged)
```

## Step 3: Clustering

```{r cluster}
merged <- cdst_cluster(merged, config)
cat("Clusters found:", length(unique(merged$cdst_cluster)), "\n")
```

## Step 4: Cell Type Annotation

```{r annotate}
merged <- cdst_annotate_cell_types(merged, config)
cat("Cell types:", length(unique(merged$cdst_cell_type)), "\n")
table(merged$cdst_cell_type)
```

### Spatial Cell Type Map

```{r spatial-ct, fig.width=12, fig.height=10}
cdst_plot_spatial(merged, color_by = "cdst_cell_type")
```

## Step 5: Neurotransmitter Classification

```{r nt}
merged <- cdst_classify_nt(merged, config)
table(merged$cdst_nt_type)
```

```{r spatial-nt, fig.width=12, fig.height=10}
cdst_plot_spatial(merged, color_by = "cdst_nt_type")
```

## Step 6: Glial Subtype Classification

```{r glia}
merged <- cdst_classify_glia(merged, config)
table(merged$cdst_glia_type, useNA = "ifany")
```

## Step 7: Cell Dynamics

```{r dynamics}
proportions <- cdst_cell_proportions(merged, config)
dist_matrix <- cdst_distribution_matrix(merged)
```

### Distribution Heatmap

```{r dist-heatmap, fig.width=14, fig.height=10}
cdst_plot_distribution_heatmap(dist_matrix)
```

## Step 8: Differential Expression

```{r deg}
deg_results <- cdst_deg_interregional(merged, config, n_cores = n_cores)
deg_summary <- cdst_summarize_deg(deg_results)
knitr::kable(head(deg_summary, 20), caption = "Top DEG Summary by Region")
```

### Volcano Plots

```{r volcano, fig.width=8, fig.height=6}
contrasts <- unique(deg_results$contrast)
for (ct in contrasts[1:min(4, length(contrasts))]) {
  p <- cdst_plot_volcano(deg_results, contrast = ct, title = ct)
  print(p)
}
```

## Step 9: WGCNA Network Analysis

```{r wgcna}
wgcna_data <- cdst_wgcna_prepare(merged, config)
wgcna_result <- cdst_wgcna_detect(wgcna_data)
trait_cor <- cdst_wgcna_trait_cor(wgcna_result)
hub_genes <- cdst_wgcna_hub_genes(wgcna_result)

knitr::kable(head(hub_genes, 20), caption = "Top Hub Genes")
```

### Module-Trait Correlations

```{r trait-heatmap, fig.width=10, fig.height=8}
cdst_plot_trait_heatmap(trait_cor, trait_type = "treatment")
```

## Step 10: Save Results

```{r save}
# Save the complete result
result <- new("CdstResult",
  config = config,
  seurat = merged,
  deg = deg_results,
  dynamics = list(proportions = proportions, distribution_matrix = dist_matrix),
  wgcna = list(wgcna_result = wgcna_result, trait_correlations = trait_cor,
               hub_genes = hub_genes),
  provenance = list(
    notebook = "01_full_pipeline.qmd",
    timestamp = as.character(Sys.time()),
    package_version = as.character(packageVersion("CellDynamicST"))
  )
)

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
saveRDS(result, file.path(output_dir, "cdst_result.rds"))
cat("Results saved to:", output_dir, "\n")
```

## Session Info

```{r session}
sessionInfo()
```
