---
title: "Cell Type Annotation Deep Dive"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cell Type Annotation Deep Dive}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

Cell type annotation in CellDynamicST follows a three-stage hierarchical
approach that mirrors the biological organization of brain tissue. This
vignette explains the methodology, parameters, and customization options for
each stage.

## The Three-Stage Annotation Pipeline

The annotation pipeline proceeds through three sequential stages, each adding
a layer of biological specificity to the cell identity assignments:

| Stage | Function | Output Column | Description |
|-------|----------|---------------|-------------|
| 1. Clustering & Annotation | `cdst_cluster()` + `cdst_annotate_cell_types()` | `cdst_cell_type` | Major cell types via reference-based RF |
| 2. NT Classification | `cdst_classify_nt()` | `cdst_nt_type` | Neurotransmitter identity (Glut, GABA, etc.) |
| 3. Glial Typing | `cdst_classify_glia()` | `cdst_glia_type` | Glial subtypes via GMM on marker expression |

## Stage 1: Reference-Based Clustering

The core of CellDynamicST's cell typing is a **reference-based Random Forest
label propagation** approach. This method was developed to handle the challenge
of annotating millions of cells from spatial transcriptomic platforms where
per-cell gene coverage is lower than scRNA-seq.

### Algorithm

The clustering and annotation process follows these steps:

1. **Dimensionality reduction**: PCA on the top variable features, followed by
   UMAP embedding for visualization.

2. **Graph-based clustering**: Shared nearest neighbor (SNN) graph construction
   followed by Louvain community detection at the configured resolution.

3. **Reference label transfer**: If a reference Seurat object is provided, a
   Random Forest classifier is trained on the reference labels and applied to
   each cluster. The majority vote across cells in each cluster determines the
   cluster's cell type assignment.

4. **Marker-based refinement**: For clusters without a confident RF assignment
   (below `rf_threshold`), marker gene expression patterns are used as a
   fallback annotation strategy.

```{r stage1}
config <- cdst_load_config("config.yaml")

# Cluster
obj <- cdst_cluster(obj, config, resolution = 0.8)

# Annotate with reference
obj <- cdst_annotate_cell_types(obj, config)

# Check results
table(obj$cdst_cell_type)
```

### Customizing Marker Genes

You can provide a custom marker gene CSV file with columns `cell_type` and
`gene`:

```{r markers}
# Custom markers
markers <- data.frame(
  cell_type = c("Excitatory", "Excitatory", "Inhibitory", "Astrocyte"),
  gene = c("Slc17a7", "Camk2a", "Gad1", "Aqp4")
)
write.csv(markers, "my_markers.csv", row.names = FALSE)
```

## Stage 2: Neurotransmitter Classification

Neurotransmitter classification assigns each neuron to its primary
neurotransmitter identity based on the expression of canonical transporter and
synthetic enzyme genes. The classification uses a criteria-based approach
defined in a YAML file.

### Default Mouse Criteria

The package ships with default criteria for mouse brain:

| NT Type | Key Markers | Classification Rule |
|---------|-------------|---------------------|
| Glutamatergic | Slc17a7, Slc17a6 | Expression > threshold in cluster |
| GABAergic | Gad1, Gad2, Slc32a1 | Expression > threshold in cluster |
| Cholinergic | Chat, Slc18a3 | Expression > threshold in cluster |
| Dopaminergic | Th, Slc6a3 | Expression > threshold in cluster |
| Serotonergic | Tph2, Slc6a4 | Expression > threshold in cluster |

```{r nt}
obj <- cdst_classify_nt(obj, config, threshold = 0.3)
table(obj$cdst_nt_type)
```

### Custom NT Criteria

Create a custom YAML file to define your own neurotransmitter criteria:

```yaml
neurotransmitter_types:
  Glutamatergic:
    markers: ["Slc17a7", "Slc17a6", "Slc17a8"]
    threshold: 0.3
    require_any: true
  GABAergic:
    markers: ["Gad1", "Gad2", "Slc32a1"]
    threshold: 0.3
    require_any: true
```

## Stage 3: Glial Subtype Classification

Glial cells are further classified into subtypes using Gaussian Mixture Models
(GMM) on the expression of canonical glial markers. This approach captures the
continuous nature of glial identity better than hard thresholding.

### Supported Glial Subtypes

| Subtype | Key Markers | Method |
|---------|-------------|--------|
| Astrocyte | Aqp4, Gfap, Aldh1l1 | GMM on marker expression |
| Oligodendrocyte | Mbp, Plp1, Mog | GMM on marker expression |
| OPC | Pdgfra, Cspg4 | GMM on marker expression |
| Microglia | Cx3cr1, P2ry12, Tmem119 | GMM on marker expression |

```{r glia}
obj <- cdst_classify_glia(obj, config)
table(obj$cdst_glia_type, useNA = "ifany")
```

## Evaluating Annotation Quality

Use `cdst_annotation_coverage()` to evaluate how well your annotations perform
at different confidence thresholds:

```{r coverage}
coverage <- cdst_annotation_coverage(
  obj,
  column_name = "cdst_cell_type",
  thresholds = seq(0, 1, by = 0.05)
)

# Plot coverage curve
library(ggplot2)
ggplot(coverage, aes(threshold, proportion_annotated)) +
  geom_line() +
  geom_point() +
  labs(title = "Annotation Coverage vs Confidence Threshold",
       x = "Confidence Threshold",
       y = "Proportion of Clusters Annotated") +
  cdst_theme_publication()
```

## Visualizing Annotations

```{r viz}
# Spatial map colored by cell type
cdst_plot_spatial(obj, color_by = "cdst_cell_type")

# Spatial map colored by NT type
cdst_plot_spatial(obj, color_by = "cdst_nt_type")

# Distribution heatmap
dist <- cdst_distribution_matrix(obj)
cdst_plot_distribution_heatmap(dist)
```
