---
title: "CellDynamicST: Complete Pipeline Walkthrough"
subtitle: "End-to-end spatial transcriptomics analysis with the Oprm1 A118G dataset"
author: "CellDynamicST Team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
    highlight: tango
    fig_width: 12
    fig_height: 8
vignette: >
  %\VignetteIndexEntry{Complete Pipeline Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 150,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

**CellDynamicST** is an integrated R package for analyzing spatial transcriptomics
data from MERFISH/CosMx experiments. It provides a complete pipeline from raw data
ingestion through quality control, cell type classification, differential expression,
co-expression network analysis (WGCNA), and publication-quality visualization.

This notebook walks through the **entire pipeline** using a subset of the
publicly available **Oprm1 A118G** spatial transcriptomics dataset
(GEO: [GSE326553](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE326553)),
which profiles ~950,000 cells across mouse brain sections from four experimental
groups in a 2×2 factorial design (Genotype: AA/GG × Treatment: SAL/MOR).

The subset included here contains **996 cells** sampled from all four slides,
focusing on major brain regions. This is sufficient to demonstrate every analysis
step and visualization function.

## Experimental Design

| Group    | Genotype | Treatment | Slides     |
|----------|----------|-----------|------------|
| AA SAL   | AA (WT)  | Saline    | Slide 1, 4 |
| AA MOR   | AA (WT)  | Morphine  | Slide 1, 4 |
| GG SAL   | GG (KI)  | Saline    | Slide 2, 3 |
| GG MOR   | GG (KI)  | Morphine  | Slide 2, 3 |

## Package Installation

```{r install, eval=FALSE}
# Install from GitHub
devtools::install_github("GoogleXie/CellDynamicST", ref = "v1.1.0")

# Or from a local tarball
install.packages("CellDynamicST_1.1.0.tar.gz", repos = NULL, type = "source")
```

# Setup and Configuration

## Load the Package

```{r load-packages}
# Source all package R files (for development; use library() for installed package)
r_files <- sort(list.files(
  file.path(system.file(package = "CellDynamicST"), "..", "CellDynamicST", "R"),
  pattern = "\\.R$", full.names = TRUE
))
# If running from source directory:
if (length(r_files) == 0) {
  r_files <- sort(list.files("R", pattern = "\\.R$", full.names = TRUE))
}
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(pheatmap)
})
for (f in r_files) {
  tryCatch(source(f, local = FALSE), error = function(e) {
    message("Note: ", basename(f), " - ", e$message)
  })
}
```

## Load the Example Dataset

The package ships with a 996-cell subset from the Oprm1 A118G experiment.
This metadata table contains all the annotations produced by the full pipeline:
cell type classifications, brain region assignments, neurotransmitter types,
and spatial coordinates.

```{r load-data}
# Load the example metadata subset
data_path <- system.file("extdata", "example_metadata_subset.csv",
                         package = "CellDynamicST")
if (data_path == "") {
  # Fallback for running from source directory
  data_path <- file.path("inst", "extdata", "example_metadata_subset.csv")
}
metadata <- read.csv(data_path, stringsAsFactors = FALSE)

cat("Dataset dimensions:", nrow(metadata), "cells x", ncol(metadata), "columns\n")
cat("Experiment groups:", paste(unique(metadata$experiment_group), collapse = ", "), "\n")
cat("Brain regions:", paste(unique(metadata$level_3_parent_roi_acronym), collapse = ", "), "\n")
cat("Cell types:", length(unique(metadata$high_level_cell_type)), "high-level types\n")
```

```{r data-summary}
# Quick summary of the experimental design
table(metadata$experiment_group, metadata$slide)
```

## Create a Configuration File

For a full analysis, you would create a YAML configuration file. Here we
demonstrate the structure:

```{r config-template, eval=FALSE}
# Generate a template configuration file
cdst_create_config_template("my_experiment_config.yaml")

# Then edit it to match your experiment:
# - Set project name and species
# - Define experimental groups (genotype × treatment)
# - Point to your Seurat .rds files or CosMx directories
# - Adjust QC thresholds and clustering parameters
```

For this walkthrough, we work directly with the pre-processed metadata,
bypassing the Seurat-dependent steps (registration, QC, clustering) and
focusing on the **analysis and visualization** modules.

# Module 1: Data Registration (Overview)
{#mod1}

In a full pipeline run, `cdst_load_seurat()` reads your Seurat objects or
CosMx directories, and `cdst_register_atlas()` maps each cell to the Allen
Brain Atlas CCFv3 coordinate system.

```{r mod1-overview, eval=FALSE}
# Full pipeline usage (requires Seurat objects):
config <- cdst_load_config("experiment_config.yaml")

# Load and merge samples
obj <- cdst_load_seurat("/path/to/slide1.rds", config)

# Register to Allen Brain Atlas
obj <- cdst_register_atlas(obj, config)
```

The registration step adds spatial coordinates (`AP_location`, `DV_location`,
`ML_location`) and brain region annotations at multiple hierarchy levels
(`level_3_parent_roi_acronym` through `level_7_parent_roi_acronym`) to each cell.

Our example subset already has these columns:

```{r mod1-check}
# Verify atlas registration columns exist
spatial_cols <- c("AP_location", "DV_location", "ML_location",
                  "roi_acronym", "level_3_parent_roi_acronym")
cat("Atlas columns present:", all(spatial_cols %in% colnames(metadata)), "\n")

# Show spatial coordinate ranges
cat("AP range:", range(metadata$AP_location, na.rm = TRUE), "\n")
cat("DV range:", range(metadata$DV_location, na.rm = TRUE), "\n")
cat("ML range:", range(metadata$ML_location, na.rm = TRUE), "\n")
```

# Module 2: Quality Control
{#mod2}

The QC module filters cells based on transcript counts, gene counts, and
negative probe rates. In a full run:

```{r mod2-overview, eval=FALSE}
# Full pipeline usage:
obj <- cdst_run_qc(obj, config)
obj <- cdst_normalize(obj, config)
obj <- cdst_find_variable_genes(obj, config)
```

## QC Visualization

Even with pre-processed data, we can visualize QC metrics:

```{r qc-plot, fig.width=14, fig.height=5}
# Add genotype and treatment columns for the QC plot
metadata$genotype <- ifelse(grepl("AA", metadata$experiment_group), "AA", "GG")
metadata$treatment <- ifelse(grepl("MOR", metadata$experiment_group), "MOR", "SAL")

cdst_plot_qc_standalone(
  data = metadata,
  group_col = "experiment_group",
  count_col = "nCount_RNA",
  feature_col = "nFeature_RNA",
  output_file = "qc_summary.png",
  width = 14, height = 5
)
```

This three-panel figure shows:

- **Left**: Distribution of transcripts per cell across experimental groups
- **Center**: Distribution of genes detected per cell
- **Right**: Scatter of genes vs. transcripts, colored by group

These plots help identify batch effects and confirm that QC thresholds are appropriate.

# Module 3: Cell Type Classification
{#mod3}

CellDynamicST uses a hierarchical classification approach:

1. **Clustering** via Louvain on a PCA-reduced neighborhood graph
2. **Neurotransmitter typing** based on marker gene expression
3. **Glial subtyping** using canonical glial markers
4. **Hierarchy annotation** building a multi-level cell type tree

```{r mod3-overview, eval=FALSE}
# Full pipeline usage:
obj <- cdst_cluster(obj, config)
obj <- cdst_classify_nt(obj, nt_col = "neurotransmitter_type")
obj <- cdst_classify_glia(obj, glia_col = "high_level_cell_type")
obj <- cdst_annotate_hierarchy(obj, config)
```

## Cell Type Distribution

```{r celltype-summary}
# High-level cell type distribution
ct_table <- table(metadata$high_level_cell_type)
ct_df <- data.frame(
  cell_type = names(ct_table),
  count = as.integer(ct_table)
) %>% arrange(desc(count))
knitr::kable(ct_df, caption = "High-level cell type counts in the subset")
```

## Cell Type Hierarchy Visualization

The hierarchy captures the relationships between classification levels:
highest-level (Neuronal/Glial/Other) → high-level cell type → neurotransmitter
type → detailed annotation.

### Alluvial / Sankey Diagram

```{r hierarchy-sankey, fig.width=16, fig.height=10}
# Build the hierarchy from metadata
hierarchy <- cdst_build_hierarchy(
  data = metadata,
  levels = c("highest_level_cell_type", "high_level_cell_type",
             "neurotransmitter_type", "cell_type_annotation"),
  min_cells = 3
)

cat("Hierarchy:", nrow(hierarchy$nodes), "nodes,",
    nrow(hierarchy$links), "links\n")

# Plot as alluvial/Sankey flow diagram
cdst_plot_hierarchy_tree(
  hierarchy_data = hierarchy,
  output_file = "hierarchy_alluvial.png",
  width = 16, height = 10
)
```

This alluvial diagram shows how cells flow from broad categories (left) to
specific annotations (right), with band widths proportional to cell counts.
Color coding distinguishes the three highest-level groups: Neuronal, Glial,
and Other Cells.

### Interactive Sankey (HTML)

```{r hierarchy-interactive, eval=requireNamespace("networkD3", quietly=TRUE)}
# Interactive version using networkD3
cdst_plot_hierarchy_sankey(
  hierarchy_data = hierarchy,
  output_file = "hierarchy_sankey.html"
)
cat("Interactive Sankey saved to hierarchy_sankey.html\n")
```

# Module 4: Cell Proportions and Dynamics
{#mod4}

## Cell Proportions Across Groups

```{r proportions, fig.width=10, fig.height=7}
cdst_plot_proportions_standalone(
  data = metadata,
  celltype_col = "high_level_cell_type",
  group_col = "experiment_group",
  output_file = "cell_proportions.png",
  width = 10, height = 7
)
```

This stacked bar chart reveals how cell type composition varies across the
four experimental groups. Differences in proportions between AA and GG genotypes,
or between SAL and MOR treatments, may indicate genotype-dependent cellular
responses to morphine.

## Disproportion Analysis

Disproportion scores quantify how cell type proportions shift between two
conditions, expressed as z-scores. Positive values indicate enrichment in
condition B; negative values indicate depletion.

```{r disproportion-calc}
# Calculate disproportion scores: AA SAL vs AA MOR
scores_aa <- cdst_disproportion_scores(
  data = metadata,
  group1 = "AA SAL",
  group2 = "AA MOR",
  group_col = "experiment_group",
  celltype_col = "high_level_cell_type",
  min_cells = 3
)

knitr::kable(
  scores_aa %>% arrange(desc(abs(zscore))) %>% head(10),
  digits = 3,
  caption = "Top disproportion scores: AA SAL vs AA MOR"
)
```

### Disproportion on t-SNE

```{r disproportion-tsne, fig.width=10, fig.height=8}
# Add tSNE coordinates (using spatial coords as proxy for this demo)
metadata$tSNE_1 <- metadata$AP_location * 3 + rnorm(nrow(metadata), 0, 0.5)
metadata$tSNE_2 <- metadata$DV_location * 3 + rnorm(nrow(metadata), 0, 0.5)

cdst_plot_disproportion(
  data = metadata,
  scores = scores_aa,
  x_col = "tSNE_1",
  y_col = "tSNE_2",
  celltype_col = "high_level_cell_type",
  title = "AA: SAL vs MOR",
  output_file = "disproportion_tsne.png",
  width = 10, height = 8
)
```

Cells are colored by their disproportion z-score: **blue** indicates enrichment
in SAL (depletion in MOR), **red** indicates enrichment in MOR. Labels at
cell type centroids identify which populations are most affected.

### Multi-Comparison Disproportion

For a 2×2 design, we typically compare both genotypes:

```{r disproportion-multi, fig.width=10, fig.height=8}
# GG SAL vs GG MOR
scores_gg <- cdst_disproportion_scores(
  data = metadata,
  group1 = "GG SAL",
  group2 = "GG MOR",
  group_col = "experiment_group",
  celltype_col = "high_level_cell_type",
  min_cells = 3
)

cdst_plot_disproportion(
  data = metadata,
  scores = scores_gg,
  x_col = "tSNE_1", y_col = "tSNE_2",
  celltype_col = "high_level_cell_type",
  title = "GG: SAL vs MOR",
  output_file = "disproportion_gg.png",
  width = 10, height = 8
)
```

Comparing the AA and GG disproportion maps reveals genotype-specific
cellular responses to morphine treatment.

# Module 5: Multi-Dimensional Heatmap
{#mod5}

The multi-dimensional heatmap is a signature visualization of CellDynamicST.
It displays cell type × brain region interactions using:

- **Circle size**: cell count
- **Circle color**: percentage change between conditions
- **Top margin**: stacked bars for cell type distribution
- **Right margin**: stacked bars for neurotransmitter composition

```{r multidim-heatmap-data}
# Prepare the heatmap input CSV
# This requires: cell_type, brain_region, group, count, pct_change columns
heatmap_data <- metadata %>%
  filter(experiment_group %in% c("AA SAL", "AA MOR")) %>%
  group_by(high_level_cell_type, level_3_parent_roi_acronym, experiment_group) %>%
  summarise(count = n(), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from = experiment_group,
    values_from = count,
    values_fill = 0
  )

# Calculate percentage change
heatmap_data$pct_change <- ifelse(
  heatmap_data$`AA SAL` > 0,
  (heatmap_data$`AA MOR` - heatmap_data$`AA SAL`) / heatmap_data$`AA SAL` * 100,
  0
)

# Write CSV for the Python-based heatmap
write.csv(heatmap_data, "heatmap_input_aa.csv", row.names = FALSE)
cat("Heatmap data:", nrow(heatmap_data), "cell type × region combinations\n")
```

```{r multidim-heatmap, eval=FALSE}
# The multi-dimensional heatmap uses a Python backend
# Requires: matplotlib, numpy, pandas
cdst_plot_multidim_heatmap(
  csv_path = "heatmap_input_aa.csv",
  output_file = "multidim_heatmap_aa.png",
  title = "AA: MOR vs SAL",
  width = 14, height = 10
)
```

# Module 6: Differential Expression Analysis
{#mod6}

## DEG Analysis (Overview)

In a full run, `cdst_deg_interregional()` performs differential expression
testing within each brain region for each pairwise comparison:

```{r mod6-overview, eval=FALSE}
# Full pipeline usage:
deg_results <- cdst_deg_interregional(
  obj, config,
  contrasts = list(
    c(group1 = "AA SAL", group2 = "AA MOR"),
    c(group1 = "GG SAL", group2 = "GG MOR")
  ),
  region_column = "level_3_parent_roi_acronym",
  min_cells_per_region = 30,
  test_method = "wilcox",
  logfc_threshold = 0.25
)
```

## Volcano Plot

The volcano plot shows the relationship between fold change and statistical
significance for differentially expressed genes.

```{r volcano, fig.width=10, fig.height=8}
# Generate example DEG results for visualization
set.seed(42)
n_genes <- 500
deg_data <- data.frame(
  gene = paste0("Gene_", seq_len(n_genes)),
  avg_log2FC = c(rnorm(200, -1.5, 0.8), rnorm(100, 0, 0.3), rnorm(200, 1.5, 0.8)),
  p_val_adj = c(
    10^(-runif(200, 2, 10)),
    10^(-runif(100, 0, 1.5)),
    10^(-runif(200, 2, 10))
  ),
  stringsAsFactors = FALSE
)

cdst_plot_volcano(
  deg_data = deg_data,
  title = "AA SAL vs MOR",
  logfc_threshold = 0.5,
  pval_threshold = 0.05,
  output_file = "volcano_aa.png",
  width = 10, height = 8
)
```

Genes are colored by significance and direction: **blue** = significantly
downregulated, **red** = significantly upregulated, **grey** = not significant.
The top genes by significance are labeled.

## GO Enrichment Comparison (2×2 Design)

The bidirectional GO enrichment plot is specifically designed for 2×2 factorial
experiments. It compares pathway enrichment between two conditions (e.g., AA vs GG)
using a weighted score that combines fold change and significance.

```{r go-enrichment, fig.width=12, fig.height=10}
# Generate example GO enrichment data for two conditions
set.seed(42)
n_terms <- 15
go_terms <- paste0("GO:", sprintf("%07d", sample(1:9999999, n_terms)),
                   " ", c(
  "synaptic transmission", "dopamine metabolic process",
  "GABA signaling pathway", "glutamate receptor signaling",
  "opioid receptor binding", "calcium ion transport",
  "axon guidance", "myelination",
  "neuroinflammation", "apoptotic process",
  "protein phosphorylation", "ion channel activity",
  "neurotransmitter transport", "cell adhesion",
  "transcription regulation"
))

enrichment_A <- data.frame(
  go_term = go_terms,
  log2FC = rnorm(n_terms, 0.8, 0.5),
  qvalue = 10^(-runif(n_terms, 1, 5)),
  n_genes = sample(5:50, n_terms, replace = TRUE),
  stringsAsFactors = FALSE
)

enrichment_B <- data.frame(
  go_term = go_terms,
  log2FC = rnorm(n_terms, -0.5, 0.6),
  qvalue = 10^(-runif(n_terms, 0.5, 4)),
  n_genes = sample(5:50, n_terms, replace = TRUE),
  stringsAsFactors = FALSE
)

cdst_plot_enrichment_lollipop(
  enrichment_A = enrichment_A,
  enrichment_B = enrichment_B,
  label_A = "AA MOR vs SAL",
  label_B = "GG MOR vs SAL",
  fc_col = "log2FC",
  pval_col = "qvalue",
  term_col = "go_term",
  n_genes_col = "n_genes",
  output_file = "go_enrichment_comparison.png",
  width = 12, height = 10
)
```

This plot shows:

- **Central bars**: Direction and magnitude of the difference between conditions
  (orange = condition A stronger, light blue = condition B stronger)
- **Left dots**: Weighted enrichment score for condition A
- **Right dots**: Weighted enrichment score for condition B
- **Dot size**: Absolute log2 fold change
- **Dot color**: Statistical significance (-log10 q-value)

This design is optimized for identifying pathways that are differentially
affected between genotypes in response to treatment.

## Enrichment Heatmap

```{r enrichment-heatmap, fig.width=10, fig.height=8}
# Generate example enrichment scores across brain regions
regions <- c("Isocortex", "HPF", "CTXsp", "STR", "HY", "MB")
enrichment_matrix <- matrix(
  rnorm(n_terms * length(regions), 0, 2),
  nrow = n_terms, ncol = length(regions),
  dimnames = list(
    gsub("GO:\\d+ ", "", go_terms),
    regions
  )
)

cdst_plot_enrichment_heatmap(
  enrichment_matrix = enrichment_matrix,
  title = "GO Enrichment Across Brain Regions",
  output_file = "enrichment_heatmap.png",
  width = 10, height = 8
)
```

The enrichment heatmap shows how GO term enrichment varies across brain
regions, with hierarchical clustering revealing regional patterns.

# Module 7: WGCNA Co-Expression Networks
{#mod7}

## WGCNA Pipeline (Overview)

Weighted Gene Co-expression Network Analysis identifies modules of
co-expressed genes and correlates them with experimental traits.

```{r mod7-overview, eval=FALSE}
# Full pipeline usage:
wgcna_data <- cdst_wgcna_prepare(obj, config, min_cells = 100)
wgcna_result <- cdst_wgcna_detect(wgcna_data, config)
trait_cor <- cdst_wgcna_trait_cor(wgcna_result)
hub_genes <- cdst_wgcna_hub_genes(wgcna_result, n_hubs = 10)
```

## Module-Trait Correlation Heatmap

```{r trait-heatmap, fig.width=8, fig.height=6}
# Generate example module-trait correlation data
modules <- c("brown", "blue", "green", "turquoise", "red",
             "yellow", "black", "pink")
traits <- c("Treatment", "Genotype", "Interaction")

trait_data <- data.frame(
  module = rep(modules, each = length(traits)),
  trait = rep(traits, times = length(modules)),
  correlation = c(
    0.82, 0.15, 0.45,   # brown
    -0.71, 0.32, -0.28,  # blue
    0.55, -0.68, 0.12,   # green
    -0.45, 0.78, -0.55,  # turquoise
    0.33, -0.22, 0.67,   # red
    -0.18, 0.45, -0.33,  # yellow
    0.62, -0.15, 0.28,   # black
    -0.55, 0.38, -0.42   # pink
  ),
  p_value = c(
    0.001, 0.35, 0.02,
    0.003, 0.08, 0.12,
    0.01, 0.001, 0.55,
    0.02, 0.001, 0.01,
    0.06, 0.22, 0.002,
    0.35, 0.02, 0.06,
    0.005, 0.42, 0.15,
    0.01, 0.04, 0.03
  ),
  stringsAsFactors = FALSE
)

cdst_plot_trait_heatmap(
  trait_data = trait_data,
  output_file = "trait_heatmap.png",
  width = 8, height = 6
)
```

The module-trait heatmap shows the Pearson correlation between each WGCNA
module eigengene and experimental traits. Significance is indicated by stars
(* p < 0.05, ** p < 0.01, *** p < 0.001). Modules with strong treatment
correlations are candidates for morphine-responsive gene programs.

## Hub Gene Network

```{r hub-network, fig.width=10, fig.height=10}
# Generate example hub gene network data
set.seed(42)
hub_genes <- paste0("Hub_", LETTERS[1:3])
neighbor_genes <- paste0("Gene_", 1:15)
all_genes <- c(hub_genes, neighbor_genes)

edges <- data.frame(
  from = c(
    rep(hub_genes[1], 6), rep(hub_genes[2], 5), rep(hub_genes[3], 4),
    sample(neighbor_genes, 8, replace = TRUE)
  ),
  to = c(
    sample(neighbor_genes, 6), sample(neighbor_genes, 5),
    sample(neighbor_genes, 4), sample(neighbor_genes, 8)
  ),
  weight = runif(23, 0.3, 1.0),
  stringsAsFactors = FALSE
)
edges <- edges[edges$from != edges$to, ]
edges <- edges[!duplicated(paste(edges$from, edges$to)), ]

nodes <- data.frame(
  gene = all_genes,
  is_hub = all_genes %in% hub_genes,
  module = "brown",
  kME = c(runif(3, 0.85, 0.99), runif(15, 0.3, 0.8)),
  stringsAsFactors = FALSE
)

cdst_plot_network(
  edges = edges,
  nodes = nodes,
  hub_genes = hub_genes,
  title = "Brown Module Hub Network",
  output_file = "hub_network.png",
  width = 10, height = 10
)
```

Hub genes (gold, larger) are the most connected nodes within a module.
Their neighbors (light blue) represent co-expressed genes that may share
regulatory mechanisms. Edge thickness reflects co-expression strength.

## WGCNA Circos Plot

```{r circos, eval=FALSE}
# The circos plot uses a Python backend
# Requires: matplotlib, numpy
cdst_plot_wgcna_circos(
  module_data_csv = "module_enrichment.csv",
  output_file = "wgcna_circos.png",
  width = 12, height = 12
)
```

The circos plot displays multiple WGCNA modules as segments around a circle,
with concentric rings showing different experimental comparisons (e.g.,
OUD enrichment, AA MOR vs SAL, GG MOR vs SAL, genotype effects).

## Atlas Network Visualization

```{r atlas-network, eval=FALSE}
# The atlas network overlays co-expression on Allen Brain Atlas contours
# Supports sagittal, coronal, and horizontal planes
cdst_plot_wgcna_network(
  nodes_csv = "network_nodes.csv",
  edges_csv = "network_edges.csv",
  output_file = "atlas_network_sagittal.png",
  plot_type = "atlas",
  plane = "sagittal",   # or "coronal" or "horizontal"
  slice = NULL,          # NULL for automatic midpoint
  width = 12, height = 8
)
```

The atlas network visualization places brain region nodes at their anatomical
coordinates on a specific brain slice, with edges colored by module membership
and weighted by co-expression strength. Users can configure the slice plane
and coordinate in the YAML configuration file.

# Module 8: Spatial Enrichment
{#mod8}

Spatial enrichment analysis maps Gene Ontology term enrichment scores onto
spatial coordinates, revealing where specific biological processes are
active or depleted.

```{r spatial-enrichment, fig.width=10, fig.height=8}
# Generate example spatial enrichment data
set.seed(42)
spatial_data <- metadata[, c("AP_location", "DV_location")]
spatial_data$enrichment_score <- rnorm(nrow(metadata), 0, 1)
# Make scores spatially structured
spatial_data$enrichment_score <- spatial_data$enrichment_score +
  0.5 * sin(spatial_data$AP_location) * cos(spatial_data$DV_location)

cdst_plot_spatial_enrichment(
  data = spatial_data,
  x_col = "AP_location",
  y_col = "DV_location",
  score_col = "enrichment_score",
  title = "Synaptic Transmission Enrichment (AA MOR)",
  subtitle = "Midbrain",
  output_file = "spatial_enrichment.png",
  width = 10, height = 8
)
```

Orange regions indicate spatial zones where the GO term is enriched;
purple regions indicate depletion. This reveals the spatial heterogeneity
of biological processes across brain tissue.

# Running the Full Pipeline
{#full-pipeline}

For a complete end-to-end analysis, create a configuration file and call
`cdst_run()`:

```{r full-pipeline, eval=FALSE}
# 1. Create and edit your configuration
cdst_create_config_template("my_config.yaml")
# Edit my_config.yaml with your experimental details...

# 2. Run the full pipeline
result <- cdst_run(
  config_path = "my_config.yaml",
  steps = c("register", "preprocess", "cluster",
            "classify_nt", "classify_glia",
            "dynamics", "deg", "wgcna", "visualize"),
  output_dir = "cdst_output",
  n_cores = 4,
  verbose = TRUE
)

# 3. View summary
cdst_summary(result)

# 4. Access individual results
seurat_obj <- result@seurat
deg_results <- result@deg
wgcna_results <- result@wgcna
```

The pipeline generates all manuscript-quality figures automatically in the
`output_dir/figures/` directory.

# Individual Visualization Functions
{#viz-reference}

All visualization functions can be called independently outside the pipeline.
Here is a quick reference:

| Function | Description | Input |
|----------|-------------|-------|
| `cdst_plot_qc_standalone()` | QC violin + scatter | metadata data.frame |
| `cdst_plot_proportions_standalone()` | Stacked bar chart | metadata data.frame |
| `cdst_plot_hierarchy_tree()` | Alluvial/Sankey flow | hierarchy list |
| `cdst_plot_hierarchy_sankey()` | Interactive Sankey (HTML) | hierarchy list |
| `cdst_plot_disproportion()` | Disproportion on t-SNE | metadata + scores |
| `cdst_plot_multidim_heatmap()` | Circle heatmap (Python) | CSV path |
| `cdst_plot_volcano()` | Volcano plot | DEG data.frame |
| `cdst_plot_enrichment_lollipop()` | Bidirectional GO comparison | Two enrichment data.frames |
| `cdst_plot_enrichment_heatmap()` | GO × region heatmap | enrichment matrix |
| `cdst_plot_spatial_enrichment()` | Spatial enrichment map | spatial data.frame |
| `cdst_plot_trait_heatmap()` | Module-trait correlation | trait data.frame |
| `cdst_plot_network()` | Hub gene network (R) | edges + nodes data.frames |
| `cdst_plot_wgcna_circos()` | Multi-ring circos (Python) | CSV path |
| `cdst_plot_wgcna_network()` | Atlas network (Python) | CSV paths + plane config |
| `cdst_plot_distribution_heatmap()` | Cell type × region heatmap | distribution matrix |
| `cdst_plot_deg_scatter()` | MA plot | DEG data.frame |

# Session Info

```{r session-info}
sessionInfo()
```
