---
title: "Configuration Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Configuration Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

CellDynamicST uses a single YAML configuration file as the source of truth for
every analysis. This design eliminates hardcoded paths, group names, and
thresholds scattered across scripts, making analyses fully reproducible and
portable across datasets.

This vignette documents every configuration field, its purpose, and valid
values.

## Configuration Structure

The configuration file has five top-level sections:

| Section | Purpose |
|---------|---------|
| `project` | Project metadata (name, organism, atlas version) |
| `samples` | Sample file paths and group assignments |
| `design` | Experimental design: groups, genotype/treatment mappings |
| `qc` | Quality control thresholds |
| `clustering` | Clustering and annotation parameters |

## Project Section

```yaml
project:
  name: "Oprm1 A118G Spatial Transcriptomic Study"
  organism: "mouse"          # "mouse" or "human"
  atlas_version: "CCFv3"     # Allen Brain Atlas version
  output_dir: "cdst_output"  # Default output directory
```

The `organism` field determines which gene annotation databases are used for GO
enrichment and pathway analysis. Currently, `mouse` (using `org.Mm.eg.db`) and
`human` (using `org.Hs.eg.db`) are supported.

## Samples Section

Each sample entry requires an `id`, a `path` to the Seurat `.rds` file, and a
`group` assignment that matches one of the groups defined in the `design`
section.

```yaml
samples:
  - id: "mouse_01_AA_SAL"
    path: "data/raw/mouse_01.rds"
    group: "AA_SAL"
    metadata:                    # Optional extra metadata
      batch: "batch_1"
      sex: "male"
  - id: "mouse_02_GG_MOR"
    path: "data/raw/mouse_02.rds"
    group: "GG_MOR"
```

The `metadata` sub-section is optional and allows you to attach any additional
sample-level information that will be propagated to the Seurat object's
metadata during registration.

## Design Section

The design section defines the factorial structure of your experiment. The
`genotype_map` and `treatment_map` dictionaries map each group name to its
genotype and treatment level, respectively.

```yaml
design:
  groups:
    - "AA_SAL"
    - "AA_MOR"
    - "GG_SAL"
    - "GG_MOR"
  genotype_map:
    AA_SAL: "AA"
    AA_MOR: "AA"
    GG_SAL: "GG"
    GG_MOR: "GG"
  treatment_map:
    AA_SAL: "SAL"
    AA_MOR: "MOR"
    GG_SAL: "SAL"
    GG_MOR: "MOR"
  contrasts:                     # Optional: explicit contrasts
    - ["AA_SAL", "GG_SAL"]
    - ["AA_MOR", "GG_MOR"]
    - ["AA_SAL", "AA_MOR"]
    - ["GG_SAL", "GG_MOR"]
```

If `contrasts` is omitted, the pipeline automatically generates all pairwise
comparisons from the `groups` list.

## QC Section

Quality control thresholds control cell filtering. All thresholds have sensible
defaults that can be overridden.

```yaml
qc:
  min_genes_per_cell: 200        # Minimum genes detected per cell
  max_genes_per_cell: 10000      # Maximum genes (doublet filter)
  min_counts_per_cell: 500       # Minimum UMI counts
  max_counts_per_cell: 100000    # Maximum UMI counts
  max_pct_mito: 20               # Maximum mitochondrial percentage
  min_cells_per_gene: 10         # Minimum cells expressing a gene
```

## Clustering Section

Parameters for dimensionality reduction, clustering, and cell type annotation.

```yaml
clustering:
  n_variable_features: 3000      # Number of highly variable genes
  n_pcs: 30                      # PCA dimensions to use
  resolution: 0.8                # Louvain clustering resolution
  algorithm: "louvain"           # "louvain", "leiden", or "slm"
  annotation:
    marker_genes_path: null      # Path to custom marker gene CSV
    reference_path: null         # Path to reference Seurat object
    rf_ntree: 500                # Random Forest trees for label transfer
    rf_threshold: 0.6            # Minimum RF probability for assignment
    nt_criteria_path: null       # Custom NT criteria YAML
    glia_criteria_path: null     # Custom glia criteria YAML
```

## Loading and Validating Configuration

```{r load}
library(CellDynamicST)

# Load and validate
config <- cdst_load_config("experiment_config.yaml")

# The config object is an S4 CdstConfig class
print(config)

# Access fields
config@project_name
config@groups
config@design$genotype_map
```

Validation checks include: all sample paths exist, all groups in samples match
the design, genotype/treatment maps cover all groups, and QC thresholds are
within valid ranges.

## Creating Configuration Programmatically

```{r programmatic}
# Start from template
template_path <- system.file("templates", "experiment_config.yaml",
                              package = "CellDynamicST")
config_list <- yaml::read_yaml(template_path)

# Modify
config_list$project$name <- "My New Experiment"
config_list$samples <- list(
  list(id = "s1", path = "data/s1.rds", group = "WT_SAL"),
  list(id = "s2", path = "data/s2.rds", group = "KO_MOR")
)

# Write
yaml::write_yaml(config_list, "my_config.yaml")
```
