---
title: "Quick Start Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

CellDynamicST is a comprehensive R package for analyzing spatial transcriptomic
data with a focus on cell dynamics, neurotransmitter classification, and gene
co-expression network analysis. The package implements an 8-module pipeline that
takes raw MERFISH/Visium/MERSCOPE data through quality control, cell type
annotation, differential expression, and weighted gene co-expression network
analysis (WGCNA).

This vignette walks through a minimal end-to-end analysis using the package's
pipeline orchestrator.

## Installation

```{r install}
# Install from GitHub
remotes::install_github("YourLab/CellDynamicST")

# Or with all suggested dependencies
remotes::install_github("YourLab/CellDynamicST", dependencies = TRUE)
```

## Step 1: Prepare Your Configuration

Every CellDynamicST analysis begins with a YAML configuration file that
describes your experimental design. Copy the template and edit it for your
experiment:

```{r config}
library(CellDynamicST)

# Copy the template to your working directory
file.copy(
  system.file("templates", "experiment_config.yaml", package = "CellDynamicST"),
  "my_experiment_config.yaml"
)
```

Edit `my_experiment_config.yaml` to specify your sample paths, experimental
groups, genotype/treatment mappings, and brain region hierarchy. The
configuration file is the single source of truth for all downstream analyses.

Here is a minimal example:

```yaml
project:
  name: "My Spatial Experiment"
  organism: "mouse"

samples:
  - id: "sample_01"
    path: "data/sample_01.rds"
    group: "WT_SAL"
  - id: "sample_02"
    path: "data/sample_02.rds"
    group: "WT_MOR"

design:
  groups: ["WT_SAL", "WT_MOR", "KO_SAL", "KO_MOR"]
  genotype_map:
    WT_SAL: "WT"
    WT_MOR: "WT"
    KO_SAL: "KO"
    KO_MOR: "KO"
  treatment_map:
    WT_SAL: "SAL"
    WT_MOR: "MOR"
    KO_SAL: "SAL"
    KO_MOR: "MOR"
```

## Step 2: Run the Full Pipeline

The simplest way to run the entire analysis is with `cdst_run()`:

```{r run}
result <- cdst_run(
  config_path = "my_experiment_config.yaml",
  output_dir = "cdst_output",
  n_cores = 4,
  verbose = TRUE
)
```

This executes all 10 pipeline steps in sequence:

1. **Register** — Load samples and standardize metadata
2. **Preprocess** — QC filtering and SCTransform normalization
3. **Cluster** — PCA, UMAP, and Louvain clustering
4. **Annotate** — Marker-based cell type annotation
5. **Classify NT** — Neurotransmitter type classification
6. **Classify Glia** — Glial subtype classification via GMM
7. **Dynamics** — Cell type proportion analysis
8. **DEG** — Interregional differential expression
9. **WGCNA** — Gene co-expression network analysis
10. **Visualize** — Generate all publication figures

Each step saves a checkpoint `.rds` file, so you can resume from any point.

## Step 3: Run Individual Steps

You can also run specific steps:

```{r steps}
# Only run preprocessing and clustering
result <- cdst_run(
  "my_experiment_config.yaml",
  steps = c("register", "preprocess", "cluster"),
  output_dir = "cdst_output"
)

# Later, resume with annotation and downstream
result <- cdst_run(
  "my_experiment_config.yaml",
  steps = c("annotate", "classify_nt", "deg", "wgcna"),
  output_dir = "cdst_output"
)
```

## Step 4: Use Individual Functions

For fine-grained control, call module functions directly:

```{r individual}
config <- cdst_load_config("my_experiment_config.yaml")

# Load and register
samples <- cdst_register_samples(config)
merged <- cdst_merge_samples(samples, config)

# QC and normalize
merged <- cdst_qc_filter(merged, config)
merged <- cdst_normalize(merged, config)

# Cluster
merged <- cdst_cluster(merged, config, resolution = 0.8)

# Annotate
merged <- cdst_annotate_cell_types(merged, config)
merged <- cdst_classify_nt(merged, config)

# Visualize
cdst_plot_spatial(merged, color_by = "cdst_cell_type")
cdst_plot_qc(merged)
```

## Step 5: Explore Results

```{r results}
# Print summary
cdst_summary(result)

# Access the annotated Seurat object
seurat_obj <- result@seurat

# Access DEG results
deg_table <- result@deg
head(deg_table)

# Access WGCNA results
hub_genes <- result@wgcna$hub_genes
head(hub_genes)

# Check provenance
cat(yaml::as.yaml(result@provenance))
```

## Output Structure

After a full pipeline run, the output directory contains:

```
cdst_output/
├── checkpoint_01_registered.rds
├── checkpoint_02_preprocessed.rds
├── checkpoint_03_clustered.rds
├── checkpoint_04_annotated.rds
├── checkpoint_05_nt_classified.rds
├── checkpoint_06_glia_classified.rds
├── checkpoint_07_dynamics.rds
├── checkpoint_08_deg.rds
├── checkpoint_09_wgcna.rds
├── cdst_result.rds
├── provenance.yaml
└── figures/
    ├── qc_summary.pdf
    ├── spatial_cell_types.pdf
    ├── distribution_heatmap.pdf
    ├── volcano_*.pdf
    └── wgcna_trait_heatmap.pdf
```

## Next Steps

For more detailed guides, see:

- `vignette("configuration")` — Full configuration reference
- `vignette("cell-typing")` — Cell type annotation deep dive
- `vignette("wgcna")` — WGCNA network analysis guide
- `vignette("visualization-gallery")` — All available plot types
