---
title: "Visualization Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

CellDynamicST provides a suite of publication-quality visualization functions
that follow a consistent design language. All plotting functions return ggplot2
objects (or pheatmap objects for heatmaps) and use the `cdst_theme_publication()`
theme for consistent styling.

This vignette showcases every available plot type with example code.

## Theme and Palette

All plots use a shared publication theme and color palette:

```{r theme}
library(CellDynamicST)
library(ggplot2)

# The publication theme
p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  cdst_theme_publication()

# Access the default palette
colors <- cdst_palette(8)
print(colors)
```

## QC Summary (`cdst_plot_qc`)

A multi-panel figure showing transcript counts, gene counts, and their
relationship across experimental groups. Essential for evaluating data quality
before downstream analysis.

```{r qc}
cdst_plot_qc(seurat_obj, group_by = "experiment_group")
```

## Spatial Cell Type Map (`cdst_plot_spatial`)

Scatter plot of cells at their spatial coordinates, colored by any metadata
column. Supports faceting by experimental group for side-by-side comparisons.

```{r spatial}
# Basic spatial map
cdst_plot_spatial(obj, color_by = "cdst_cell_type")

# Faceted by group
cdst_plot_spatial(obj, color_by = "cdst_nt_type",
                  facet_by = "experiment_group",
                  point_size = 0.2)

# Custom palette
my_colors <- c("Glutamatergic" = "#E41A1C", "GABAergic" = "#377EB8",
               "Cholinergic" = "#4DAF4A", "Dopaminergic" = "#984EA3")
cdst_plot_spatial(obj, color_by = "cdst_nt_type", palette = my_colors)
```

## Cell Type Distribution Heatmap (`cdst_plot_distribution_heatmap`)

Heatmap showing the proportion of each cell type across brain regions, with
hierarchical annotations for parent regions and cell type superclasses.

```{r dist_heatmap}
dist <- cdst_distribution_matrix(obj, normalize = "row")
cdst_plot_distribution_heatmap(dist,
  color_scale = "viridis",
  title = "Cell Type Distribution Across Brain Regions")
```

## Volcano Plot (`cdst_plot_volcano`)

Standard volcano plot for differential expression results, with configurable
significance thresholds and automatic labeling of top genes.

```{r volcano}
cdst_plot_volcano(deg_results,
  contrast = "WT_SAL vs WT_MOR",
  p_threshold = 0.05,
  logfc_threshold = 0.5,
  n_label = 20)
```

## DEG Scatter with Displacement (`cdst_plot_deg_scatter`)

Scatter plot showing DEGs per brain region with gene-level points displaced
by fold change. Supports functional group coloring for biological
interpretation.

```{r deg_scatter}
functional_groups <- list(
  Synaptic = c("Syp", "Stxbp1", "Gabra1"),
  Metabolism = c("App", "Pfkm", "Ldhb"),
  Ion_Transport = c("Atp2a2", "Slc8a1", "Calb1")
)

cdst_plot_deg_scatter(deg_results,
  functional_groups = functional_groups,
  title = "DEGs by Region and Functional Group")
```

## WGCNA Module-Trait Heatmap (`cdst_plot_trait_heatmap`)

Correlation heatmap between WGCNA module eigengenes and experimental traits,
with significance stars and diverging color scale.

```{r trait_heatmap}
cdst_plot_trait_heatmap(trait_cor, trait_type = "treatment")
cdst_plot_trait_heatmap(trait_cor, trait_type = "genotype")
cdst_plot_trait_heatmap(trait_cor, trait_type = "region")
```

## Gene Co-expression Network (`cdst_plot_network`)

Interactive network visualization of gene co-expression, with nodes colored
by WGCNA module and hub genes highlighted with larger size and labels.

```{r network}
# Single module
cdst_plot_network(wgcna_result,
  hub_genes = hub_genes,
  module = "blue",
  min_cor = 0.3)

# All modules
cdst_plot_network(wgcna_result,
  hub_genes = hub_genes,
  min_cor = 0.4)
```

## Spatial Enrichment Heatmap (`cdst_plot_spatial_enrichment`)

Heatmap of GO-term enrichment scores across brain regions, showing which
biological processes are spatially enriched or depleted between conditions.

```{r spatial_enrich}
cdst_plot_spatial_enrichment(enrichment_results,
  contrast = "WT_SAL vs WT_MOR",
  title = "Spatial GO Enrichment: Treatment Effect")
```

## Saving Figures

All ggplot2-based plots can be saved with `ggsave()`. For pheatmap objects,
wrap in a PDF device:

```{r save}
# ggplot2 plots
p <- cdst_plot_volcano(deg_results)
ggsave("figures/volcano.pdf", p, width = 8, height = 6, dpi = 300)

# pheatmap plots
pdf("figures/distribution_heatmap.pdf", width = 14, height = 10)
cdst_plot_distribution_heatmap(dist)
dev.off()
```

## Customizing the Theme

Override specific theme elements while keeping the base publication style:

```{r customize}
p <- cdst_plot_spatial(obj) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom",
    legend.text = element_text(size = 8)
  )
```
