---
title: "WGCNA Network Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WGCNA Network Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

CellDynamicST implements an **interregional** WGCNA workflow that identifies
gene co-expression modules across brain regions and experimental conditions.
Unlike standard WGCNA applied to individual cells, this approach aggregates
expression at the region-group level, treating each unique brain region ×
experimental group combination as a sample. This captures spatial patterns of
gene co-regulation that are invisible at the single-cell level.

## The WGCNA Workflow

The complete WGCNA workflow consists of five sequential steps:

| Step | Function | Purpose |
|------|----------|---------|
| 1. Prepare | `cdst_wgcna_prepare()` | Build region × gene matrix |
| 2. Detect | `cdst_wgcna_detect()` | Module identification |
| 3. Correlate | `cdst_wgcna_trait_cor()` | Module-trait associations |
| 4. Hub Genes | `cdst_wgcna_hub_genes()` | Identify hub genes per module |
| 5. Preserve | `cdst_wgcna_preservation()` | Cross-group preservation |

An optional sixth step performs pathway enrichment on each module.

## Step 1: Prepare the Expression Matrix

The preparation step constructs a region-by-gene average expression matrix.
Each row represents a unique combination of brain region and experimental
group (e.g., "CA1_WT_SAL"), and each column represents a gene. Regions with
fewer than `min_cells` cells are excluded to ensure reliable averages.

```{r prepare}
library(CellDynamicST)

config <- cdst_load_config("config.yaml")
wgcna_data <- cdst_wgcna_prepare(obj, config, min_cells = 100)

# Inspect dimensions
dim(wgcna_data$datExpr)  # rows = region-groups, cols = genes
head(wgcna_data$traits_df)
```

## Step 2: Module Detection

Module detection uses the standard WGCNA blockwiseModules algorithm with
automatic soft threshold selection. The function returns the complete network
object, module assignments, and module eigengenes.

```{r detect}
wgcna_result <- cdst_wgcna_detect(
  wgcna_data,
  power = NULL,           # Auto-select (recommended)
  min_module_size = 20,
  deep_split = 3,
  merge_cut_height = 0.10
)

# Module summary
table(wgcna_result$module_colors)
```

### Manual Power Selection

If automatic power selection fails or you want to inspect the scale-free
topology fit, you can examine the soft threshold analysis:

```{r power}
sft <- wgcna_result$sft

# Plot scale-free topology fit
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit (signed R²)",
     type = "n")
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = sft$fitIndices[, 1], cex = 0.9, col = "red")
abline(h = 0.90, col = "red", lty = 2)

# Re-run with manual power
wgcna_result <- cdst_wgcna_detect(wgcna_data, power = 8)
```

## Step 3: Module-Trait Correlations

This step computes Pearson correlations between module eigengenes and
experimental traits (genotype, treatment, brain region). The resulting
correlation and p-value matrices are used for heatmap visualization.

```{r trait}
trait_cor <- cdst_wgcna_trait_cor(wgcna_result)

# Visualize
cdst_plot_trait_heatmap(trait_cor, trait_type = "treatment")
cdst_plot_trait_heatmap(trait_cor, trait_type = "genotype")
cdst_plot_trait_heatmap(trait_cor, trait_type = "region")
```

## Step 4: Hub Gene Identification

Hub genes are the most highly connected genes within each module, identified
by their module membership (kME) score. These genes are the strongest
candidates for driving the biological processes captured by each module.

```{r hubs}
hub_genes <- cdst_wgcna_hub_genes(
  wgcna_result,
  n_hubs = 10,
  kme_threshold = 0.7
)

head(hub_genes)

# Visualize network with hub genes highlighted
cdst_plot_network(wgcna_result, hub_genes = hub_genes, module = "blue")
```

## Step 5: Module Preservation

Module preservation analysis tests whether modules detected in one
experimental group are conserved in other groups. High preservation indicates
robust, biologically meaningful modules; low preservation may indicate
condition-specific gene regulation.

```{r preserve}
preservation <- cdst_wgcna_preservation(
  wgcna_result,
  n_permutations = 200
)
```

## Optional: Pathway Enrichment

Perform GO biological process enrichment on each module to understand the
biological functions captured by the co-expression patterns.

```{r enrich}
enrichment <- cdst_wgcna_pathway_enrich(
  wgcna_result,
  organism = "mouse",
  p_cutoff = 0.05
)

head(enrichment)
```

## Complete Example

```{r complete}
library(CellDynamicST)

config <- cdst_load_config("config.yaml")

# Full WGCNA pipeline
wgcna_data   <- cdst_wgcna_prepare(obj, config)
wgcna_result <- cdst_wgcna_detect(wgcna_data)
trait_cor    <- cdst_wgcna_trait_cor(wgcna_result)
hub_genes    <- cdst_wgcna_hub_genes(wgcna_result)
enrichment   <- cdst_wgcna_pathway_enrich(wgcna_result)

# Generate all WGCNA figures
cdst_plot_trait_heatmap(trait_cor, trait_type = "treatment")
cdst_plot_trait_heatmap(trait_cor, trait_type = "genotype")
cdst_plot_network(wgcna_result, hub_genes = hub_genes, module = "blue")
```
